openarchaeoCollaboration

### 2023-06-14 analysis/_04-what_is_open_archaeology.qmd
Add main document openarchaeo_collaboration.qmd

Each record is annotated with tags that describe the aspects of archaeological work that each tool contributes to. Items are tagged based on how they have been generally identified or presented by their maintainers. The list of tags and their associated definitions is subject to grow or change as archaeologists continue to develop software for additional use cases. Right now there are 63 tags.

Records are also categorized based on how each tool or resource is accessed or used. This gives us a sense of the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold. In other words, categories hint at some of the decisions that developers make regarding how they want their tools to be used and accessed.

@@

We defined two kinds of categories: software and documents. Software comprise collections of instructions that dictate how a computer should operate, and documents are bounded sets of recorded information. We define three categories of software and four categories of documents. Packages and libraries are sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform. Platforms are the broader systems upon which packages or libraries are plugged into or are meant to extend upon, such as R, Python, QGIS, and Meshlab, which each have their own protocols to facilitate such integration. Standalone software are tools that may be operated without needing to first access an underlying platform. The third category of software, scripts, are sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation. Things like code generalized sets of code snippets.

As for documents, specifications, protocols and schemas are formal data structures or frameworks intended to be used as models. Products are stable outcomes of creative work, loosely defined. Guides are educational resources or documented protocols meant to instruct readers how to apply relevant tools or techniques. And finally, lists or datasets, including open-archaeo itself, are series of consistently organized observations assembled with purpose.

@@

Our first set of analyses were focused on these questions:

What kinds of software are archaeologists making?

And what are the general purposes of their software?

Most projects included in open-archaeo are designed to be used as a package or library atop am existing 'platform' -- for example a package for a programming language like R, or a plugin for an application like QGIS. The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes. Others create standalone software that can be run independently of such platforms, for example desktop or web apps. A significant amount of projects also comprise of datasets and non-packaged code snippets that have been made available for general use.

@@

R is overwhelmingly dominant as a platform, closely followed by Python. Plugins for QGIS and mobile apps are also relatively popular. Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS); a number of lesser used programming languages; and a genre consisting of custom forms and spreadsheet templates. In the figure below, the darker grey lines are used to break up the bars by primary author, showing that many of the smaller platforms---for example Excel templates---are targeted by only one or two developers. The larger platforms tend to be more diverse.

@@

If we compare categories with tags, we can see the general domains that each kind of resource is designed to serve. We see that packages are fairly common across the board. Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling. These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

@@

---

- Add in a breakdown on licensing practices here too

>>>>

We annotated each record with tags that describe aspects of archaeological work that each tool contributes to. We assigned tags based on how their maintainers identified their projects. The list of tags and their associated definitions is subject to grow or change as archaeologists continue to develop software for additional use cases. Right now there are 63 tags.

We also categorized records based on how each tool or resource is meant to be accessed or used. This gives us a sense of the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold. In other words, categories hint at some of the decisions that developers make regarding how they want their tools to be used and accessed.

@@

We defined two kinds of categories: software and documents. Software comprise collections of instructions that dictate how a computer should operate, and documents are bounded sets of recorded information. We define three categories of software and four categories of documents. Packages and libraries are sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform. Platforms are the broader systems upon which packages or libraries are plugged into or are meant to extend upon, such as R, Python, QGIS, and Meshlab, which each have their own protocols to facilitate such integration. Standalone software are tools that users may operate without needing to first access an underlying platform. The third category of software, scripts, are sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.

The first kind of document, which we refer to as specifications, protocols and schemas, are formal data structures or frameworks intended to be used as models. Products are stable outcomes of creative work, loosely defined. Guides are educational resources or documented protocols meant to instruct readers how to apply relevant tools or techniques. And finally, lists or datasets, including open-archaeo itself, are series of consistently organized observations assembled with purpose.

@@

Our first set of analyses focused on these questions:

* What kinds of software are archaeologists making?
* What are the general purposes of their software?

Most projects included in open-archaeo are designed to be used as a package or library atop am existing "platform" -- for example a package for a programming language like R, or a plugin for an application like QGIS. The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes. Others create standalone software that can be run independently of such platforms, for example desktop or web apps. A significant amount of projects also comprise of datasets and non-packaged code snippets that have been made available for general use.

@@

R is overwhelmingly dominant as a platform, closely followed by Python. Plugins for QGIS and mobile apps are also relatively popular. Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates. In the figure below, the darker grey lines are used to break up the bars by primary author, showing that many of the smaller platforms---for example Excel templates---are targeted by only one or two developers. The larger platforms tend to be more diverse.

@@

When we compare categories with tags, we see the general domains that each kind of resource is designed to serve. We see that packages are fairly common across the board. Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling. These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

@@

### 2023-06-14 analysis/_07-network.qmd
Add main document openarchaeo_collaboration.qmd

This graph identifies three core clusters characterized by repositories whose contributors commit to projects other than their own. The clusters that appear along the peiphery largely correspond with the work of single individuals, and sometimes also their colleagues. Peripheral clusters that are connected to core clusters by only a few relationships represent the sole (and perhaps initial) integrations of lone developers into a broader community.^[**NOTE:** I wonder what these repositories that prompt greater integration are? They should correspond with the nodes of highest betweenness centrality among those in the core cluster.]^[**NOTE:** I also wonder how recently these connections were forged, or how they are situated within users' history of commits.]

The three core clusters also have distinct character. One is characterized by its focus on archaeogenetics, which consists of a very well established collaborative network and general reliance on data modelling and processing tools. A second cluster is mostly centred on fieldwork-oriented data collection tools, and particularly the emergence of dominant platforms that attract more attention than other independent projects scattered across the network. The third and most significant cluster includes a schmorgasborg of projects whose contributors share varied interests. The emphasis in this latter cluster is on the formation of a central software development community rather than on any specific topic of work.^[**NOTE:** How does this map onto real-world research centres or partnerships? I'm wary of getting too into the weeds on specific people or relationships, but we might highlight the fact that projects led by people who are more active in the SSLA are well represented in this cluster, as well as members of ISAA-Kiel. Note that this observation is not backed by statistics, it is just a casual remark based on me browsing the network visualization.]

@@

We assembled graphs linking users based on common contributions to the same repositories. We sought to identify whether certain users, who contribute in certain distinct ways, play different roles in overall network. We found that the people with the highest betweenness values are those who primarily produce computational archaeology code as their job.

**NOTE:** Create a graph visualization showing either edge-betweenness clustering or rendering nodes' relative betweenness by size.
**NOTE:** We can further investigate this point made above by comparing with some other measures of activity, like volume of commits.

We applied the same betweenness centrality algorithm on user-user networks whose links are based on only commits and on only issues and comments. For the latter, people who commit less code are more represented with higher betweenness scores, but many of the people with high betweenness from the graph representing all contributions also appear here. These people who appear in both of these lists have a tendency to contribute as both committers and as commenters.

@@

@@

* What proportion of users and repos are connected to this network?
* What predicts membership in this network?
* Does centrality in the network correlate with general rate of activity?
*  Who are the most central nodes in this network? Is there anything they have in common?
   * High activity?
   * Certain platforms/categories/languages?
* Qualitative analysis of top 50? 100? most connected users
  * Career stage? Precarity? Employer? Location?

In cleaning the data we made a few significant decisions that revealed insights into our dataset. We constructed the user-user graph and filtered out several bots. We filtered out several large non-archaeology projects that effected the network in ways that distracted us from what we wanted to say about the archaeological community. We filtered isolates. We calculated the degree and betweenness to assess whether this was a realistic representation of what we know. We wanted to figure out who are the key nodes, what they had in common.

The people who are high betweenness are people who primarily produce computational archaeology code as a job. We can investgate further by comparing with some other measures of activity, like volume of commits or something else. We also looked at the same thing, but with only commits, or without commits. For the subgraph of non-commits, it includes many people from the commits-only or commits-included graphs, but also more people who do not commit as much.

>>>>

This graph identifies three core clusters characterized by repositories whose contributors commit to projects other than their own. The clusters that appear along the peiphery largely correspond with the work of single individuals, and sometimes also their colleagues. Peripheral clusters that are connected to core clusters by only a few relationships represent the sole (and perhaps initial) integrations of lone developers into a broader community.

**NOTE:** Create a table containing the nodes with highest betweenness centrality from the repo-repo graph. Columns should specify the name of the repository, the number of commits, the category, and tags. If possible also include a summary of the pulse as a line chart with x axis as time and y axis as activity. This will allow us to identify the characteristics of these repositories that prompt greater integration.

The three core clusters have their own distinct character. One has a focus on archaeogenetics, which consists of a very well established collaborative network and general reliance on data modelling and processing tools. A second cluster is mostly centred on fieldwork-oriented data collection tools, and particularly the emergence of well-funded and well-supported dominant platforms that attract more attention than other independent projects scattered across the network. The third and most significant cluster includes a schmorgasborg of projects whose contributors share varied interests. The emphasis in this latter cluster is on the formation of a central software development community rather than on any specific topic of work. Many of the projects represented in this third cluster emerge from underlying professional partnerships, namely research labs (e.g. ISAA-Kiel) and special interest groups (SSLA).

@@

We assembled graphs linking users based on common contributions to the same repositories.

**NOTE:** Plot a graph that renders nodes' sizes based on their betweenness centrality.

@@

We sought to identify whether certain users, who contribute in certain distinct ways, play different roles in overall network. We found that the people with the highest betweenness values are those who primarily produce computational archaeology code as their job. Moreover, we found that these people tend to be employed under precarious circumstances. Although precarious employment is part of our sad reality, in the context of developing and maintaining open source software, this presents a serious source of risk. The people who occupy central positions in these networks are crucial community members that make the network whole, and if they are either unable to continue on in their contributions or decide to leave archaeology entirely, then the overall network would fragment.

**NOTE:** Create a table containing the top 10 users based on betweenness centrality values, with columns specifying theur username, real name, job title/affiliation, location, career stage (student, early/mid/late career, tenured/non-tenured, independent, etc).

We also applied the same betweenness centrality algorithm on a subnetwork whose links are based only on issues and comments, and not code contributions. In this subnetwork, people who commit less code have higher betweenness scores. However, many of the people with high betweenness from the graph representing all contributions also appear here. These people who appear in both of these lists have a tendency to contribute as both committers and as commenters. This list also includes a series of contributors who never or rarely commit code. Although it is out of this study's scope a qualitative analysis of issues and comments may yield more insight on the kinds of contributions that each of these participants make.

**NOTE:** Create a table containing the top 10 users based on betweenness centrality values derived from the issues and comments only subnetwork, with columns specifying theur username, real name, job title/affiliation.

@@
